---
title: "PDF version of exercises"
output: 
  html_document:
    toc: false
---

```{r, echo=FALSE, purl = FALSE}
library(knitr)
library(stringr)
knitr::opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=FALSE)
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(stringr))
source('reveal.R')
```

Click on the links below to download pdf versions of the exercises. 

\ 

```{r, reveal, echo=FALSE, eval = isFALSE(show_exercise)}
cat("Exercises will be released as the course progresses")
```

### Day 1

\  

 - [Graphical data exploration exercise](./exercises/graphical_data_exploration_exercise.pdf)

### Day 2

\  

 - [Linear model with single continuous explanatory variable exercise](./exercises/linear_model_1_exercise.pdf)

### Day 3

\  

 - [Linear model with single categorical explanatory variable exercise](./exercises/linear_model_2_exercise.pdf)
 
 - [Linear model with categorical and continuous explanatory variables](./exercises/linear_model_3_exercise.pdf)

### Day 4

\  

- [Linear model selection exercise](./exercises/linear_model_4_exercise.pdf)